Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-08 paper/paper_draft.qmd
Updates for tracked changes to rest of methods, results and discussion

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For scenario assessment, coral cover time series for each
model run were scaled to represent cover as a percentage of the total estimated reef area.
The scenario-median time series for each reef was used for further analyses. Time series of
coral cover were then clustered based on their behaviour. Assessment metrics used
in analyses can be found in @stbl-metrics.

@@

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted at three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were
conducted at the different spatial scales as, conceptually, the temporal behaviour
constituting "low", "medium" or "high" coral cover may differ at different spatial scales.
As it was, analyses at the whole-of-GBR scale were found to lead to identical conclusions
regarding reef characteristics which influence coral cover and so for brevity are not shown
here.

@@

Partial Dependence Analysis (PDA) was conducted to understand relationships between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, outgoing connectivity, and log10-transformed
outgoing-to-incoming connectivity ratio. Conceptually, PDA is related to a Regional
Sensitivity Analysis [@hornberger1981; @pianosi2016], as model results are assessed to map
the influence of X - the predictors or model factors - on Y - the modelled outcomes
according to a metric of interest.

@@

Individual feature analysis for carrying capacity indicates that its influence on cluster
assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This
aligns with the right-skewed distribution of carrying capacity across reefs for these three
predictor variables. Beyond this low-end range, partial dependence becomes near constant,
indicating outlier values for these factors are not influential. Assessing random forest
prediction accuracy across bioregions, reveals increasing performance as mean bioregion reef
depth increases (@sfig-rf-predictor-perf). This highlights that unexplained variation in the
test dataset (33% of variation) may be concentrated in shallow bioregions where available
model predictors do not capture differences between clustered reefs.

@@

This investigation's results show increasing reef depth has a strong positive impact on
modelled reef resilience in temporal behaviour and maintenance of positive carbonate budgets
under possible future conditions. Reefs with a median depth of 10m or greater, with high
incoming connectivity, are the most likely to be assigned to a high cover time series
cluster. By itself, incoming connectivity alone does not correlate with maintenance of
positive carbonate budgets. These findings highlight the importance of depth and incoming
connectivity as coupled selection features when assessing potential reef resilience under
future thermal stress.

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While reefs experience similar
magnitude DHWs based on SST, the proportion of a reefs' coral population impacted by heat
stress events depends on reef depth. Results of this investigation suggest that for
non-mesophotic reef ecosystems (i.e., \< 30m), depth contributes greatly to reef resilience
and maintenance positive carbonate budgets under future conditions. Movement of shallow
reefs towards negative carbonate budgets in the future may contribute to phase shifts in
reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to maintain
coral driven CaCO~3~ production and accumulation into the future for longer. Spatially these
results indicate that regions of positive carbonate budget maintenance, under ACCESS-ESM1-5,
and high depth levels were concentrated in the Far Northern, offshore Townsville/Whitsunday,
and Mackay/Capricorn management areas (@sfig-depth-map; @sfig-carb-budget-map). Resilient
reefs in the Far Northern management area were exposed to lower mean DHW levels than other
reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however similar spatial patterns were
observed across GCMs for reefs clustered into the high cover cluster
(@sfig-gcm-cluster-map). Spatial findings are not interpreted as accurate predictions of
future resilience due to high uncertainty in input data, particularly depth and
connectivity. While similar findings have been demonstrated in the Pacific [@smith2014;
@baird2018] and Indo-Pacific [@bridge2014] regions, depth refugia analyses have often
focused on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific.

The results of this investigation support the hypothesis that reefs at greater depths will
exhibit characteristics of refugia and prolonged resilience (in terms of coral cover) under
climate change compared to shallower reefs. Additionally, Liu, Callaghan and Baldock
[-@liu2024] suggest that rubble instability is inversely correlated with reef depth,
possibly limiting juvenile coral recruitment capacity of shallow reefs. A caveat to
depth-resilience findings is the quality of the bathymetry data used to calculate median
reef depth. As detailed previously, estimates of median reef depth are derived from
satellite imagery (at $10m^2$ resolution) which can be subject to high degrees of
uncertainty, both in terms of the estimated depth for each pixel as well as their alignment
to the established GBRMPA reef outlines.

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. No direct, first order, relationship between weighted incoming connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions was
found. There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes, with findings dependent on the larval connectivity model used. These
investigations were performed using prior generations of larval connectivity models
parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods, without
considering the influence of reef depth on resilience. Using updated 1km^2^ resolution
larval connectivity across the whole GBR, with a limited temporal sample of particle flows,
it was found that larval connectivity alone may not be important or reef resilience and
recovery, but should be considered in combination with more influential factors such as
depth.

>>>>

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@lecuyer2018; @kucherenko2015]. For scenario assessment, coral cover time
series for each model run were scaled to represent cover as a percentage of the total
estimated reef area. The scenario-median time series for each reef was used for further
analyses. Time series of coral cover were then clustered based on their behaviour.
Assessment metrics used in analyses can be found in @stbl-metrics.

@@

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted at three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were
conducted at the different spatial scales as, conceptually, the temporal behaviour
constituting "low", "medium" or "high" coral cover may differ at different spatial scales.
Analyses at the whole-of-GBR scale were found to lead to identical conclusions regarding
reef characteristics which influence coral cover and so for brevity are not shown here.

@@

Partial Dependence Analysis (PDA) was conducted to understand relationships between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, outgoing connectivity, and log10-transformed
outgoing-to-incoming connectivity ratio. Conceptually, PDA is related to a Regional
Sensitivity Analysis [@hornberger1981; @pianosi2016], as model results are assessed to map
the influence of $X$ - the predictors or model factors - on $y$ - the modelled outcomes
according to a metric of interest.

@@

Individual feature analysis for carrying capacity indicates that its influence on cluster
assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This
aligns with the right-skewed distribution of carrying capacity across reefs for these three
predictor variables. Beyond this low-end range, partial dependence becomes near constant,
indicating outlier values for these factors are not influential. Assessing random forest
prediction accuracy across bioregions, reveals increasing performance as mean bioregion reef
depth increases (@sfig-rf-predictor-perf). Unexplained variation in the test dataset (33% of
variation) may be concentrated in shallow bioregions where considered model predictors do
not capture differences between clustered reefs.

@@

Results show increasing reef depth has a strong positive impact on modelled reef resilience
in temporal behaviour and maintenance of positive carbonate budgets under possible future
conditions. Reefs with a median depth of 10m or greater, with high incoming connectivity,
are the most likely to be assigned to a high cover time series cluster. By itself, incoming
connectivity alone does not correlate with maintenance of positive carbonate budgets. These
findings highlight the importance of depth and incoming connectivity as coupled selection
features when assessing potential reef resilience under future thermal stress.

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels
experienced and exhibit overall similar DHW time series. While reefs experience similar
magnitude DHWs based on SST, the proportion of a reefs' coral population impacted by heat
stress events depends on reef depth. Results of this investigation suggest that for
non-mesophotic reef ecosystems (i.e., \< 30m), depth contributes greatly to reef resilience
and maintenance positive carbonate budgets under future conditions.

Movement of shallow reefs towards negative carbonate budgets in the future may contribute to
phase shifts in dominant sources of CaCO~3~ production. Reefs may switch away from
coral-driven processes, and CaCO~3~ accretion, to encrusting algae driven CaCO~3~
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008]. Reefs at increased depths may be able to maintain
coral driven CaCO~3~ production and accumulation into the future for longer.

Spatially, these results indicate that regions of positive carbonate budget maintenance,
under ACCESS-ESM1-5, and high depth levels were concentrated in the Far Northern, offshore
Townsville/Whitsunday, and Mackay/Capricorn management areas (@sfig-depth-map;
@sfig-carb-budget-map). Resilient reefs in the Far Northern management area were exposed to
lower mean DHW levels than other reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however
similar spatial patterns were observed across GCMs for reefs clustered into the high cover
cluster (@sfig-gcm-cluster-map).

These findings are not to be interpreted as accurate predictions of future resilience due to
high uncertainty in input data, particularly depth and connectivity. While similar findings
have been demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific
[@bridge2014] regions, depth refugia analyses have often focused on the potential of
mesophotic reefs (\> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015;
@semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between
depth and the alleviation of heat stress across a range of shallow reefs in the Pacific.

This investigation supports the hypothesis that reefs at greater depths will exhibit
characteristics of refugia and prolonged resilience (in terms of coral cover) under climate
change compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024]
suggest that rubble instability is inversely correlated with reef depth, possibly limiting
juvenile coral recruitment capacity of shallow reefs. A caveat to depth-resilience findings
is the quality of the bathymetry data used to calculate median reef depth. As detailed
previously, estimates of median reef depth are derived from satellite imagery (at 10m^2^
resolution) which can be subject to high degrees of uncertainty, both in terms of the
estimated depth for each pixel as well as their alignment to the established GBRMPA reef
outlines.

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008;
@torres2024]. No direct, first order, relationship between weighted incoming connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions was
found. There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes, with findings dependent on the larval connectivity model used. These
investigations were performed using prior generations of larval connectivity models
parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods, without
considering the influence of reef depth on resilience. Using updated 1km^2^ resolution
larval connectivity across the whole GBR, with a limited temporal sample of particle flows,
it was found that larval connectivity alone may not be important or reef resilience and
recovery but should be considered in combination with more influential factors such as
depth.
